% ==============================================================================
% CHAPTER 3: SELF-PROTECTING SYSTEMS ARCHITECTURE
% ==============================================================================

\chapter{Self-Protecting Systems Architecture}
\label{chap:sps_architecture}

This chapter details the architectural core of the proposed blockchain-based Self-Protecting System (SPS), focusing on how to address the centralized autonomic manager's problem explored in ~\ref{sec:tech_state_of_art}. The model presented herein is grounded in the foundational decentralized architecture introduced by ~\cite{11126595} and subsequently refined through the requirement-driven analysis by our research ~\cite{albero2026blockchain}. The primary objective is to formalize the system decomposition, state model, end-to-end protection workflow, threat assumptions, and resilience properties. These components justify the integration of a permissioned, Byzantine Fault Tolerant (BFT) distributed ledger as the secure decision substrate for the autonomic manager.

\section{Design Principles and Rationale}
\label{sec:design_principles}

Traditional Self-Protecting Systems are typically designed around a centralized MAPE-K (Monitor-Analyze-Plan-Execute over Shared Knowledge) feedback loop~\cite{kephart2003vision}. While functionally effective, this centralized model introduces a fundamental security asymmetry: the managed system can be replicated, segmented, and protected, whereas the autonomic manager itself remains a high-value, centralized target. If an adversary successfully compromises the manager, they can suppress valid security alerts, trigger unauthorized response actions, or stall mitigation entirely. This effectively neutralizes the self-protection mechanism and grants the attacker complete control over the production environment~\cite{chess2003security, pekaric2023systematic}.

To address this structural vulnerability, the architecture proposed in this thesis distributes the decision authority of the autonomic manager across replicated, isolated nodes coordinated through a permissioned blockchain using BFT consensus~\cite{11126595}. The core design choice is not simply the adoption of blockchain as a modern storage medium, but its use as a replicated state machine where:
\begin{itemize}
    \item \textbf{Consensus-driven state updates}: Monitored variables representing the managed system state are updated only after achieving quorum-based agreement among independent observers,
    \item \textbf{Deterministic transitions}: Response decisions are modeled as deterministic functions of the globally agreed state, guaranteeing that honest nodes reach identical conclusions,
    \item \textbf{Auditable actuation}: Actuation instructions are pushed from a tamper-evident, globally replicated log, ensuring source accountability and post-incident forensic auditability.
\end{itemize}

By encoding the autonomic manager within a Byzantine Fault Tolerant ledger, trust is shifted from a single, hardening-dependent process to a formal fault model. The central security question shifts to determining the number of components that can fail or be corrupted before the safety and liveness of the control loop are violated, exposing the managed system to threats. 

At the architectural level, this model is deliberately technology-agnostic. It does not dictate a specific Intrusion Detection System (IDS) engine, response planner, or blockchain implementation. Instead, it defines a strict security contract among the components: authenticated observations, deterministic state transitions, replicated BFT consensus, and auditable, blind actuation. A crucial aspect of this design is the requirement for \emph{immediate finality} in the underlying consensus protocol. Unlike public permissionless blockchains that rely on probabilistic finality (e.g., waiting for multiple block confirmations), a software control loop cannot act on decisions that might be subsequently rolled back. A block commit must represent a final, immutable transition to avoid executing conflicting or unauthorized actions.

\section{Architectural Decomposition and Trust Boundaries}
\label{sec:architecture_decomposition}

To ensure structural resilience, the architecture decomposes the system into distinct functional planes, separated by explicit trust boundaries and network isolation. The system maintains the classic autonomic split between the \emph{managed system} (the production environment) and the \emph{manager system} (the security controller)~\cite{kephart2003vision, 11126595}, but introduces strict boundaries to limit the lateral propagation of attacks.

\subsection{Managed System}
\label{subsec:managed_system}

The managed system represents the production workload to be protected (e.g., containerized application services, microservices, or database nodes). From a security perspective, this domain is assumed to be untrusted, as it is directly exposed to external users and potential attackers. The behavior and operational health of the managed system are captured through a set of monitored variables:
\[
\mathbf{V} = \{v_1, v_2, \dots, v_k\}.
\]
These variables represent resource-level, network-level, or application-level signals (such as CPU saturation, database query anomalies, authentication failure rates, or protocol misuse indicators). At any time instant $t$, the observable state of the managed system is represented by the vector:
\[
s_t = \langle v_1(t), v_2(t), \dots, v_k(t)\rangle.
\]
Security-relevant events (e.g., a Denial of Service attack or a SQL injection attempt) modify one or more coordinates of $s_t$. The architecture does not couple itself to any specific detection paradigm, allowing the use of signature-based, anomaly-based, or hybrid detection systems to monitor these variables.

\subsection{Manager System}
\label{subsec:manager_system}

The manager system is the actuation plane in charge of processing state observations and executing responses. It is decomposed into three logical planes:
\begin{enumerate}
    \item \textbf{Detection Plane}: Composed of multiple independent IDS instances that observe overlapping portions of the managed system's state variables $\mathbf{V}$.
    \item \textbf{Consensus-and-Decision Plane}: A permissioned blockchain network that maintains the replicated decision state and executes the response selection logic.
    \item \textbf{Actuation Plane}: A set of response executors (actuators) that receive committed mitigation commands and enforce them on the managed system.
\end{enumerate}

Each active component in the detection and actuation planes is paired with a blockchain-facing node. In our design, we distinguish between \emph{validating nodes} (which participate in the BFT consensus protocol) and \emph{member or light nodes} (which submit transactions or listen to events without validating blocks). This separation decouples functional roles from consensus overhead, allowing the validator set to remain small and efficient while the number of sensors and actuators scales.

\subsection{Isolation and Containerization Requirements}
\label{subsec:isolation_primitive}

The security of the manager system relies heavily on containerization and network isolation~\cite{11126595}. Each IDS, actuator, and blockchain node must run in its own isolated container, deployed across physical or virtual hosts. In our prototype, this environment is emulated using the Kathará framework~\cite{kathara} on top of Docker. 

Under this model, the container boundary serves as the primary security perimeter. If an attacker compromises an application service in the managed system, they cannot pivot into the manager system's containers unless an explicit, authenticated communication path exists. The attack surface of the manager system is restricted to three channels:
\begin{itemize}
    \item The telemetry and input channels of the IDSs,
    \item The transaction submission APIs exposed by the blockchain nodes,
    \item The control paths through which actuators apply countermeasures.
\end{itemize}

\subsection{Actuator Blindness and Unidirectional Control}
\label{subsec:blind_actuators}

To prevent attackers from manipulating responses, actuators are modeled as \emph{blind}~\cite{11126595}. Actuators do not query the managed system directly to inspect its state, nor do they accept instructions from it. Instead, they operate exclusively as passive consumers of the committed decisions stored in the blockchain. 

If an actuator were allowed to pull state variables directly from a compromised application container to confirm an attack, the attacker could exploit this interactive channel to inject malicious payloads or trigger unauthorized actions. By enforcing actuator blindness, the control loop establishes a strict unidirectional data flow: IDSs push observations to the consensus plane, the blockchain validates and commits the state transition, and the resulting response is pushed to the actuators. This design pattern ensures that an actuator only executes commands that have passed through BFT consensus.

\section{System's State Formalization}
\label{sec:formal_state_model}

To prevent ambiguity and support formal verification, the control pipeline of the decentralized manager is formalized as a deterministic state machine.

\subsection{Observation Sets and Redundancy}
\label{subsec:observation_sets}

Let $\mathcal{D} = \{d_1, d_2, \dots, d_m\}$ be the set of deployed IDSs. Each IDS $d_i$ is configured to monitor a specific subset of state variables, denoted by the observation scope $O_i \subseteq \mathbf{V}$. For any monitored variable $v_j \in \mathbf{V}$, we define its observer set $\mathcal{D}_{v_j}$ as the subset of IDSs that monitor it:
\[
\mathcal{D}_{v_j} = \{d_i \in \mathcal{D} \mid v_j \in O_i\}.
\]

To ensure semantic robustness, the architecture requires $|\mathcal{D}_{v_j}| \ge 3$ for all critical variables, utilizing heterogeneous detection technologies (e.g., combining Snort, Suricata, and Zeek). Using diverse detection engines ensures that a zero-day exploit or signature bypass targeting one specific IDS implementation will not compromise the consolidated observation, as the remaining heterogeneous observers can still report the event accurately.

\subsection{Proposal and Consolidation Semantics}
\label{subsec:proposal_semantics}

When an IDS $d_i$ detects a change in a variable $v_j$, it submits a state proposal transaction to the blockchain:
\[
p = \langle id_{ids}, v_j, \hat{x}, t \rangle
\]
where $id_{ids}$ is the unique identifier of the submitting IDS ($d_i$), $v_j$ is the target variable, $\hat{x}$ is the proposed value or classification, and $t$ is the detection timestamp.

Inside the smart contract or the application state machine, incoming proposals are collected within a sliding aggregation window. To filter out noise or a compromised sensor's false reports, the state is updated only when a quorum predicate is satisfied. In its simplest majority form, the consolidation predicate is defined as:
\[
count_{v_j}(\hat{x}) \ge q_j, \quad q_j = \left\lfloor\frac{|\mathcal{D}_{v_j}|}{2}\right\rfloor + 1.
\]
In stricter Byzantine configurations with $3f_d + 1$ observers, the quorum threshold can be adjusted to $2f_d + 1$ to tolerate up to $f_d$ Byzantine (malicious or failed) IDSs. 


\subsection{Replicated Decision State}
\label{subsec:replicated_decision_state}

The global state of the decentralized manager is formalized as:
\[
\Sigma_t = \langle S_t, R_t, I_t \rangle
\]
where:
\begin{itemize}
    \item $S_t$ is the consolidated monitored state (represented as a parameter-status map),
    \item $R_t$ is the response relation (represented as a state-action map mapping consolidated states to countermeasures),
    \item $I_t$ is the internal control metadata (such as active response flags, execution status, and cooldown timers).
\end{itemize}

Every committed transaction deterministically transforms the manager state: $\Sigma_t \xrightarrow{T_i} \Sigma_{t+1}$. 
To represent this state in a concrete implementation (such as the prototypes analyzed in Chapters~\ref{chap:requirements} and \ref{chap:implementation}), the consolidated monitored state $S_t$ is modeled as a bit-vector (a sequence of boolean flags) of length $N$, where $N$ represents the number of distinct attacks or anomalies detectable by the system. The response relation $R_t$ is implemented as a map associating each state vector with a specific mitigation action. For example, in our prototype, $R_t$ maps state vectors to command strings or script identifiers that the actuators are authorized to execute.

\subsection{Determinism and Auditability}
\label{subsec:determinism_auditability}

Deterministic execution is a fundamental requirement for the manager state machine. Given the same initial state $\Sigma_0$ and an identical sequence of transactions $T_1, T_2, \dots, T_n$, all non-faulty validator nodes must compute the exact same sequence of states $\Sigma_1, \dots, \Sigma_n$ and produce the exact same outputs. This provides two key security benefits:
\begin{itemize}
    \item \textbf{Consistency}: It prevents split-brain scenarios, ensuring that all replicas agree on which response action to trigger,
    \item \textbf{Forensics}: The entire history of committed transitions is recorded in an immutable ledger, allowing post-incident reconstruction and verification of security policies.
\end{itemize}

To preserve determinism across the system, both the state transition function (executed by the consensus nodes) and the alert generation process (performed by the detection components) must adhere to these constraints. For instance, any runtime logic or actuator selection mechanism must avoid relying on non-deterministic variables (such as local clocks or system randomness) so that all replicas compute identical state updates.

\subsection{State-Action Map Limitations and Temporal Context}
\label{subsec:assumption}

A direct state-action map assumes a Markov-like property, meaning that the current consolidated state $S_t$ contains all the information necessary to select a response. While this simplifies the design and guarantees deterministic replay, it limits the manager's ability to detect multi-stage or low-and-slow attacks that unfold over time. 

To overcome this limitation, the state representation must incorporate historical context directly into the state variables. This can be achieved by storing memory variables (such as event counters, sliding-window registers, and cooldown flags) within the internal metadata $I_t$. Rather than relying on external logs, the state machine itself tracks temporal patterns, ensuring that the inputs to the response logic remain self-contained and deterministic.

\section{End-to-End Protection Workflow}
\label{sec:protection_workflow}

The proposed architecture establishes a closed-loop control pipeline that processes security events from detection to mitigation. The workflow consists of four sequential, causally ordered stages:

\subsection{Managed Monitoring and Attack Detection}
\label{subsec:stage_detection}

When a malicious or anomalous event occurs in the managed system, it alters one or more monitored variables. The independent IDSs observing these variables detect the anomaly and evaluate the event type. Because different IDSs utilize different detection engines (e.g., signature-based vs. protocol analysis) and run on distinct containers, they may flag the attack at slightly different times, introducing a temporal skew. The architecture accommodates this asynchrony by aggregating alerts incrementally on the blockchain's smart contract.
\subsection{Alert and Transaction Submission}
\label{subsec:stage_submission}

The detecting IDS sends an alert to an associated Member Node that wraps the observation into a proposal, signs it cryptographically, and submits it as a transaction to its paired blockchain node. The entire permissioned blockchain network receives the transaction and verifies the signature against the membership service. Because the blockchain nodes are isolated from the untrusted managed system, an external attacker who has compromised an application container cannot easily inject forged transactions or manipulate the consensus membership without compromising multiple IDSs. This process is discussed in more detail in Section~\ref{subsec:two_layer_fault_tolerance}.

\subsection{Consensus-Guided State Transition}
\label{subsec:consensus_exec}

The validating nodes receive the verified transactions, order them using the BFT consensus protocol, and package them into blocks. Once a block is committed, the transitions are executed by the smart contract (or the ABCI application). If the aggregated proposals satisfy the voting predicate, the consolidated state $S_t$ is updated, and the response logic determines the appropriate countermeasure $a_t$. This BFT-guided commitment guarantees that the response decision is final and agreed upon by a qualified majority of the network.

\subsection{Response Actuation and Feedback Loop}
\label{subsec:stage_actuation}

Once a response $a_t$ is committed, the blockchain node notifies the selected actuator using a push-based subscription channel (e.g., Ethereum event logs or CometBFT Websocket events). The blind actuator receives the instruction and executes the mitigation action (e.g., blocking an IP address, updating firewall rules, or restarting a container) via a secure channel such as SSH. 

The enforcement of the countermeasure alters the managed system's state, returning the monitored variables $\mathbf{V}$ to their normal, secure range. The IDSs observe this return to a safe state and submit proposals to clear the active threat state. The state machine then updates, clears the active response flag, and terminates the response execution, closing the feedback loop.

\section{Correctness and Actuation Properties}
\label{sec:act_plane_properties}

To evaluate the reliability of the actuation plane, we define formal correctness properties that must be preserved by the architecture.

\subsection{Safety Invariants}
\label{subsec:act_safety}

Safety in the actuation plane requires that the manager never triggers unauthorized or conflicting actions. We express this through two formal invariants:

\begin{enumerate}
    \item \textbf{No Unauthorized Actuation}: Let $A$ be the set of mitigation actions, and let $E_{exec}(a)$ denote the event that action $a \in A$ is executed by an actuator. Let $\Sigma_{commit}(a)$ denote the event that the replicated state machine commits a transition emitting the response action $a$. The safety invariant requires:
    \[
    E_{exec}(a) \implies \Sigma_{commit}(a)
    \]
    This is guaranteed by actuator blindness and push-based communication channels.
    \item \textbf{Agreement (No Conflicting Decisions)}: For any block height $h$, and any two non-faulty consensus nodes $i$ and $j$, their committed states must be identical:
    \[
    \Sigma^i_h = \Sigma^j_h
    \]
    This is guaranteed at the blockchain layer by BFT consensus protocols, but must be properly addressed in the smart contract implementation for IDS-level voting.
\end{enumerate}

The combination of these two invariants is critical. Reaching consensus without actuator discipline (blindness) could still allow unauthorized local execution, while actuator discipline without consensus could result in divergent control histories across different nodes.

\subsection{Liveness Guarantees}
\label{subsec:control_liveness}

Liveness ensures that the system reacts to threats in a timely manner. Formally, if a security breach changes the monitored variables $v_j$ at time $t$ such that the response logic requires action $a$, then a non-faulty actuator must execute $a$ at some time $t' > t$, assuming that the number of Byzantine failures does not exceed the consensus protocol's threshold.

Liveness is practically constrained by validator availability, transaction submission latency, and block production times. The block production model of the blockchain backend is a relevant parameter when choosing the technology to implement the system and will be addressed in Chapter. //TODO aggiungere al capitolo su comparazione!!
Under a fixed-interval block production model (such as in GoQuorum), transactions must wait in the mempool for the next block timer, introducing a systematic delay. In contrast, an event-driven block production model (such as in CometBFT) creates blocks as soon as transactions are available, minimizing reaction latency under sparse threat conditions.


\section{Failure, Recovery and Security}
\label{sec:failure_recovery_semantics}

A resilient SPS must maintain predictable behavior not only during active cyber-attacks but also during system faults, partitions, and recovery events.

\subsection{Crash and Restart of Non-Validator Components}
\label{subsec:crash_restart_nonvalidators}

If an IDS container crashes, the system's detection coverage temporarily decreases, but the consensus and decision plane continues to operate. Upon restart, the IDS re-authenticates and resumes submitting proposals without requiring a system-wide reconfiguration thanks to the self-healing properties of containerized applications. 

If an actuator container crashes and restarts, it must not replay stale mitigation actions. To prevent this, the actuator queries the internal metadata $I_t$ on the blockchain. If the metadata indicates that the action associated with the current state has already been successfully executed, the actuator suppresses execution, preventing duplicate actuation cycles that could disrupt the managed system.

\subsection{Validator Failure and Quorum Loss}
\label{subsec:validator_failure_quorum}

Let $N_v$ be the total number of validator nodes in the permissioned network. In a standard BFT consensus protocol, the quorum size $Q_v$ required to commit a block is:
\[
Q_v = \left\lfloor\frac{2N_v}{3}\right\rfloor + 1.
\]
The system can tolerate up to $f_v = \lfloor \frac{N_v - 1}{3} \rfloor$ Byzantine (malicious or crashed) validators. 

If the number of failed validators remains within this threshold ($f \le f_v$), the network continues to commit transactions and guarantees safety and liveness. If the number of failures exceeds $f_v$, the consensus protocol stalls (safety-preserving stall). While the manager ceases to process new state updates, it preserves its last committed state, avoiding split-brain or divergent state transitions.

\subsection{Network Partitions}
\label{subsec:network_partitions}

In the event of a network partition, the CAP theorem applies //TODO Citare il CAP theorem, forse scriverlo in breve da qualche parte, da valutare. The BFT-based manager prioritizes consistency (safety) over availability (liveness). 
Only the partition containing a supermajority of validators ($Q_v$) can continue committing state transitions. Minority partitions are unable to reach consensus and stall. While this temporarily reduces responsiveness in isolated segments, it prevents the creation of divergent control histories, preserving the integrity of the global state.


\subsection{Threat Model}
\label{subsec:security_objectives}
The threat model assumes an active adversary with network access to the managed service, whose objective is to subvert the autonomic manager. A successful attack on the manager is defined by two failure modes:
\begin{enumerate}
    \item \textbf{Omission}: An attack occurs and is detected, but the manager fails to execute the required response,
    \item \textbf{Illegitimate Actuation}: An actuator executes a countermeasure that was not authorized by the consolidated manager state.
\end{enumerate}

The threat model is defined under the following assumptions:
\begin{itemize}
    \item The container isolation primitive is robust, preventing container escapes,
    \item Network segments connecting the manager (IDSs and Blockchain) containers are isolated from public traffic,
    \item The only direct, untrusted inputs entering the manager are the exposed variables monitored by the IDSs (e.g., the network traffic),
    \item The smart contract or ABCI state machine code is correctly and securely implemented, and
    \item Node identities are verified and governed by a permissioned membership model.
\end{itemize}

\subsection{Adversarial Strategies}
\label{subsec:adversarial_strategies}

An attacker attempting to subvert the manager can pursue three main strategies:
\begin{enumerate}
    \item \textbf{Input Poisoning}: The attacker attempts to compromise multiple IDSs to inject false observations, forcing the manager into an incorrect state.
    \item \textbf{Consensus Subversion}: The attacker attempts to compromise more than $1/3$ of the validator nodes to take control of block commitment and rewrite history.
    \item \textbf{Actuator Corruption}: The attacker attempts to compromise the actuators directly to block mitigations or trigger unauthorized actions.
\end{enumerate}

The architecture raises the cost of each strategy:
\begin{itemize}
    \item Input poisoning is mitigated by the multi-IDS voting logic and the use of heterogeneous detection engines that prevent technology-specific faults,
    \item Consensus subversion is prevented by BFT consensus, which requires compromising a supermajority of validators, and
    \item Actuator corruption is mitigated by actuator blindness, which prevents direct communication from compromised application containers.
\end{itemize}


\subsection{Voting and Validation Differences}
\label{subsec:two_layer_fault_tolerance}
This section formalizes the security guarantees of the architecture, showing how the remaining attack paths are restricted and highly resource-intensive for an adversary.
Let $N_v$ be the number of validators in the consensus layer. The system can tolerate up to $f_v$ Byzantine validators, where:
\[
f_v = \left\lfloor \frac{N_v - 1}{3} \right\rfloor.
\]
Let $N_d$ be the number of independent IDSs monitoring a state variable $v_j$. If majority voting is applied, the observation layer can tolerate up to $f_d$ compromised IDSs, where:
\[
f_d = \left\lfloor \frac{N_d - 1}{2} \right\rfloor.
\]

To successfully subvert the manager, an attacker must compromise either a majority of the IDSs observing a target variable ($> f_d$) or a supermajority of the consensus validators ($> f_v$). This two-layer threshold design prevents the single point of failure inherent in centralized autonomic managers.

Redundancy without diversity is vulnerable to common-mode failures. If all IDSs run the exact same software, a single vulnerability could allow an attacker to compromise the entire set simultaneously. The architecture therefore recommends using heterogeneous IDS engines (e.g., Snort, Suricata, and Zeek). This increases the attacker's costs by requiring them to develop and deploy multiple distinct exploits, making the possibility of compromising the autonomic manager highly improbable since the architecture can increase the number of diverse technologies at will.

\subsection{Residual Risk}
\label{subsec:residual_risk}

The architecture does not protect against host-level compromises or vulnerabilities in the underlying container runtime. If an attacker gains root access to the physical host, they can bypass container boundaries and compromise the nodes. Additionally, bugs in the smart contract or ABCI code remain a critical vulnerability, as a replicated state machine will faithfully execute and replicate flawed logic. In any computer system, expanding the trusted computing base is a choice that must be treated with caution, therefore, secure coding and formal verification of the response transition function remain essential.